Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-10-17 boettiger.tex
a few edits

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model-adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here: it has been shown that these statistics correspond to particular assumptions about 
an underlying process, and systems that violate these assumptions exhibit sudden transitions without any warning~\cite{Hastings2010}.
Elsewhere sudden transitions have been shown in ecological models where the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
All are based around a bifurcation process in which the principle eigenvalue passes through zero.

% remove skew, CV

It may appear that our approach is more limited as the models are explcitly linear with Gaussian noise processes.
This is consistent

While

Moreover, this illustrates that the summary statistics are very unreliable indicators. 
A low false alarm rate is only achieved by poor sensitivity; 
or conversely sensitivity it only achieved with a high rate of false alarms.  
This is under assumptions that are highly favorable to these statistics, corresponding directly to the process assumptions.  
The reason for this comes largely from the difficulty in quantifying an increase without a baseline -- 
the distribution of tau under the null model covers almost the entire range.  %Ref FIGURE

Our approach of estimating uncertainty and using the likelihood indicator is neither more nor less general than the other indicators we discuss.  
All rely on the assumption of an underlying bifurcation driven by the principle eigenvalue passing through zero.\footnote{
Skew, a higher order term, has additional assumption about the asymmetry,
and is thus generically less powerful than signals on the lower order moments.} 
These methods are robust to the assumptions of Gaussian noise and linearity of the model,
as illustrated by the fit to the simulated data, 
which instead comes from a nonlinear, discrete-state Markov process (see Appendix).  
The likelihood approach still performs the best on this data, hence it is not less general.

When the underlying bifurcation model does not apply, neither summary statistics nor our model-based description will be useful.

We stress that all of the early warning signals approaches must only be applied when the underlying assumptions can be justified.  
The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
and 
Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,

If we can use increasing variance as an indicator, than we can use the approach described here to quantify uncertainty.

Although not all sudden collapses correspond to our model assumptions,
including dynamical systems where the variance actually decreases as a sudden shift approaches,⁠
using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
but would be amenable to the approach used here.

@@

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig3}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated dataset has 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1, and additional examples, Fig. S2.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig4}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.} 
  \end{center}
 \end{figure*}

>>>>

By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
Other literature has already begun to address the additional challenges posed when the underlying 
dynamics do not correspond to these models~\cite{Hastings2010}.
Our results illustrate that even in this best-case scenario, 
reliable identification of warning signals from summary statistics can be difficult.  
The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
or even the distribution of that statstic under the null model of a stable system, 
but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
By estimating the ROC curve for a given set of data, 
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
We can develop indicators that get the most from this data.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model-adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
and systems that violate these assumptions can exhibit sudden transitions without any warning,
such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
Sudden transitions have also been observed in ecological models in which 
the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
As the bifurcation results from the principle eigenvalue passing through zero, 
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.  
The preformance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
(introduced by the Poisson proccess demographic events), 
demonstrates the accuracy under violation of these assumptions.

%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
%but would be amenable to the approach used here.

@@

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig3.pdf}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated dataset has 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1, and additional examples, Fig. S2.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig4.pdf}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.} 
  \end{center}
 \end{figure*}

### 2011-10-17 pnas_appendix.tex
a few edits

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

@@

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS2.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data.}
  \label{fig:S2}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS3.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
  \label{fig:S3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS4.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation I data.}
  \label{fig:S4}
\end{figure}

The ROC curve indicates the trade off between false alarms and failed detections, 
from which we may choose a threshold value that will allow us to classify the data set as having detected a warning or not.  
For instance, in the case of the \emph{Daphnia} data, the 90\% true-positive, 35\% false-positive rate corresponds with a threshold of 
$\delta = 2.5$.  The observed $\delta=6.0$, clearly above that threshold.  
We can of ask what is the probability of a false alarm or missed event at the observed threshold of 6 as well: \%9.5 false alarm rate.  
While it is possible to classify

which can guide the classification of a particular system as 
When the curve lies very close to the 1:1 line, there is essentially no information provided by the given warning indicator,
and classifying

To determine whether the data should be considered

>>>>

%\documentclass[authoryear,5p]{elsarticle}
%\documentclass[authoryear,review,11pt]{elsarticle}
\documentclass[11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

@@

\section{Distributions of the indicator statistic}
The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.  The value of the statistic observed in the original data is indicated by a black triangle. Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.  Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.  The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS2.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system.}
  \label{fig:S2}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS3.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability.}
  \label{fig:S3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS4.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
  \label{fig:S4}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS5.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data.}
  \label{fig:S5}
\end{figure}

\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
they specify two model forms (stable and changing stability)
but with model parameters that must be first estimated from the data.
Comparing models whose parameters have been estimated by maximum likelihood is first treated by~\citet{Cox1961, Cox1962},
and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.

Let $L_0$ be the likelihood function for model 0, 
let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
The statistic we will use is $\delta$, 
defined to be twice the difference in log likelihood of observing the data under the two MLE models,
$\delta = -2 (\log L_0 - \log L_1 )$.  
This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.

\pagebreak

\section{On a hypothesis-testing framework}\label{Dakos}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.  In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.  By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.

The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).  We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.  For this reason we generate the null distribution directly from the model as outlined above.   The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).  The plots in Fig. 3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.  Our approach replicate these findings for the $\tau$-statistic.  We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.

The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.  As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.  Fig. S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.  The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.  The reality is worse, we find it overlaps almost completely.  We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.

The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''

\section{R package tutorial}\label{R}
We provide an R package with simple implementation of the methods described here.  The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.  The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood-based analysis:

\begin{verbatim}
> require(warningsignals)
>
> # Load a sample dataset
> data(glaciationIII)
>
> # Fit constant (OU) model and a stability-loss model (LSN);
> models <- fit_models(glaciationIII, 'LSN')
>
> results <- montecarlotest(models$const, models$timedep, n=2000, cpu=16)
> plot(results)
\end{verbatim}

The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.  The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coefficient of variation).  Copies of data used in the paper are provided as examples.  The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.   Full documentation of functions and data are provided in the package, as well as example scripts which replicate the analyses presented in the paper.
